---
title: "Covid-19 Data Analysis"
author: "Alexander Kahanek"
date: "3/22/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


gathered from
https://data.humdata.org/dataset/novel-coronavirus-2019-ncov-cases

Data is cummulative

```{r, echo = FALSE}
library(plyr)
library(ggplot2)
library(tidyverse)
library(scales)
library(plotly)

#Function for adding a row for each specific date value, then taking those new rows...
combine_dates <- function(data, status, country){
  data = data[,c("Status", "Value", "Date", "Keep", "Country")]
  for (d in unique(data$Date)){
    #sum dates
    count = sum(data[data$Date == d,"Value"])
  
    #add rows for total dates
    data = rbind(data, list(status,count, d, "y", country))
  }
  #grab rows where we keep, and the cols we want
  data = data[data$Keep == "y",c("Status", "Value", "Date", "Country")]
  return (data) #send back
}

#comining 3 datasets
combine_3 <- function(x,y,z){
  data = rbind(x,y)
  data = rbind(data,z)
  return (data)
}

#Function for seperating countries into a specific dataframe, function all combines all dates for us
seperate_country <- function(country, df_c, df_r, df_d){
  
  #taking countries
  df1 = df_c[df_c[,"Country.Region"]== country,]
  df2 = df_r[df_r[,"Country.Region"]== country,]
  df3 = df_d[df_d[,"Country.Region"]== country,]
  
  #Adding Country row, to rename
  df1$Country = country
  df2$Country = country
  df3$Country = country
  
  #calling function on confirmed to combine dates
  df1 = combine_dates(df1,"Confirmed", country)
  df2 = combine_dates(df2,"Recovered", country)
  df3 = combine_dates(df3,"Deceased", country)
  
  #calling function to combine all 3 datasets
  data = combine_3(df1, df2, df3)
  
  #Changing dates from strings to dates
  data$Date = as.Date(data$Date, "%m/%d/%Y")
  
  return(data)
}


#Function used for taking merged dataframe, and extracting the data based on status
extract_status <- function(data, status){
  data = data[data$Status == status, names(data)]
}

#Function for converting Cummulative day to per day
cummulative_to_per <- function(data, cumm_column, p_d_column, status){
  
  #get status sent(need to get efficient ordering)
  data = extract_status(data, status)
  
  #reverse order of dates
  data = data[rev(order(as.Date(data$Date, format   = "%Y-%m-%d"))),]
  
  data[p_d_column] = 0 #set new column to 0
  
  previous_total = 0 #start at 0
  for (d in data$Date){
    
    #get current total, will be lowest to highest
    total = data[data$Date == d, "Value"]
    
    #formula for a current day
    per_day = previous_total - total
    
    #Grabbing the change rate
    rate_change = (per_day/previous_total)
    
    if (previous_total == 0){
      rate_change = NA
    }
    
    #reset previous total to loop  
    previous_total = total
    
    if (per_day == max(data$Value)*-1){
      per_day = NA #this value is unkown
    }
    
    #assign column value
    data[data$Date == d, p_d_column] = per_day
    data[data$Date == d, "Rate_Of_Change"] = rate_change
  } 
    
  return(data) #return data
}

#Function used for calling so get the cummulative data, needed to remerge the data. needed if 
get_cummulative <- function(data, cumm_column, p_d_column){
  #get data for confirmed
  data_c <- cummulative_to_per(data,cumm_column, p_d_column, "Confirmed")
  #get data for recovered  
  data_r <- cummulative_to_per(data,cumm_column, p_d_column, "Recovered")
  #get data for deceased
  data_d <- cummulative_to_per(data,cumm_column, p_d_column, "Deceased")
  
  #metge data back
  data = combine_3(data_c, data_r, data_d)
  
  #return data
  return (data)
}


#Function used for getting the totals for confirmed, recovered, deceased, and active.
get_totals <- function(data){
  countries = unique(data$Country)
  data1 <- data.frame("Country" = "none", "Total" = 0, "Status" = "none", "Percent_Global" = 0.0, "keep"="n")

  
  for (country in countries){
    confirmed <- max(data[data$Country == country & data$Status == "Confirmed","Value"])
    recovered <- max(data[data$Country == country & data$Status == "Recovered","Value"])
    deaths <- max(data[data$Country == country & data$Status == "Deceased","Value"])
    active <- confirmed - recovered - deaths
    
    max_con = max(data[data$Country == "Global" & data$Status == "Confirmed", "Value"])
    max_rec = max(data[data$Country == "Global" & data$Status == "Recovered", "Value"])
    max_dea = max(data[data$Country == "Global" & data$Status == "Deceased", "Value"])
    max_act = max_con - max_rec - max_dea
     
    #Creating new dataframes
    x_c = data.frame( "Country"= country, "Total" = confirmed, "Status" = "Confirmed", "Percent_Global" = confirmed/max_con, "keep"="y")
    x_r = data.frame( "Country"= country, "Total" = recovered, "Status" = "Recovered", "Percent_Global" = recovered/max_rec, "keep"="y")
    x_d = data.frame( "Country"= country, "Total" = deaths, "Status" = "Deceased", "Percent_Global" = deaths/max_dea, "keep"="y")
    #create a dataframe for active cases #confirmed - recovered - deaths
    x_a = data.frame( "Country"= country, "Total" = active, "Status" = "Active", "Percent_Global" = active/max_act, "keep"="y")
    
    
    #combining dataframes
    data1 = rbind(data1,x_c)
    data1 = rbind(data1,x_r)
    data1 = rbind(data1,x_d)
    data1 = rbind(data1,x_a)
    
  }
  data1 = data1[data1$Country != "none", c("Country", "Total", "Status", "Percent_Global")]
  return (data1) #sending dataframe back
}
```

# Cleaning data

Here I am doing a few things:
1. I made a row for each status
2. I got rid of all columns except Country, Status, Value, Date
3. I added a new region called "Global", this holds all the values for every country combined
4. I ran a short algorithm to get the per day change, aswell as the rate of change per day
5. I got a dataframe of the totals for each Country and Status, aswell as add the active status for this dataframe

What I need to do:
1. Build algorithm to get real-time Active status
2. Import data of testing

You can see the functions I use in the above codeblock.

```{r, echo = FALSE}
#Reading in our  three datasets
df_c <- read.csv("3-22-20-time_series-ncov-Confirmed.csv")
df_r <- read.csv("3-22-20-time_series-ncov-Recovered.csv")
df_d <- read.csv("3-22-20-time_series-ncov-Deaths.csv")

#removing all values that are 0
df_c <- df_c[df_c[,"Value"]!=0,]
df_r <- df_r[df_r[,"Value"]!=0,]
df_d <- df_d[df_d[,"Value"]!=0,]

#Adding a status
df_c$Status = "Confirmed"
df_r$Status = "Recovered"
df_d$Status = "Deceased"

#adding new columns for use later
df_c$Keep = "n"
df_r$Keep = "n"
df_d$Keep = "n"

#getting rid of other columns
df_c <- df_c[,c("Country.Region", "Status", "Value", "Date", "Keep")]
df_r <- df_r[,c("Country.Region", "Status", "Value", "Date", "Keep")]
df_d <- df_d[,c("Country.Region", "Status", "Value", "Date", "Keep")]

#Change dates to strings
df_c$Date = as.character(df_c$Date)
df_r$Date = as.character(df_r$Date) 
df_d$Date = as.character(df_d$Date)

#Lets go ahead and grab a global dataset aswell
global_c <- df_c
global_r <- df_r
global_d <- df_d

global_c$Country.Region = "Global"
global_r$Country.Region = "Global"
global_d$Country.Region = "Global"

global <- seperate_country("Global", global_c, global_r, global_d)

global <- get_cummulative(global, "Value", "Per_Day")

#Sepperating countries from all our dataframes, this will also combine all the dates for us
data_us <- seperate_country("US", df_c, df_r, df_d)
data_ch <- seperate_country("China", df_c, df_r, df_d)
data_it <- seperate_country("Italy", df_c, df_r, df_d)
data_ge <- seperate_country("Germany", df_c, df_r, df_d)
data_ir <- seperate_country("Iran", df_c, df_r, df_d)
data_fr <- seperate_country("France", df_c, df_r, df_d)

#grabbing per day values
data_us <- get_cummulative(data_us, "Value", "Per_Day")
data_ch <- get_cummulative(data_ch, "Value", "Per_Day")
data_it <- get_cummulative(data_it, "Value", "Per_Day")
data_ge <- get_cummulative(data_ge, "Value", "Per_Day")
data_ir <- get_cummulative(data_ir, "Value", "Per_Day")
data_fr <- get_cummulative(data_fr, "Value", "Per_Day")

#Combining all data into one big dataframe
data1 <- combine_3(data_us, data_ch, data_it)
data2 <- combine_3(data_ge, data_ir, data_fr)

#Almost our final dataframe, including all values
data <- rbind(data1, data2)
data <- rbind(data, global)

#Seperating our cleaned data into cases again
data_c <- extract_status(data, "Confirmed")
data_r <- extract_status(data, "Recovered")
data_d <- extract_status(data, "Deceased")


#Variable for deleting left over data variables
to_delete <- c("data_ch", "data_it", "data_ge", "data_ir", "data_fr", "data1", "data2", "df_c", "df_r", "df_d", "global", "global_c", "global_d", "global_r", "to_delete")

#removing old variables
rm(list = to_delete)

#Countries I want to compare
countries <- c("China", "Italy", "US")

#Get totals dataset
totals <- get_totals(data)
```

```{r}
head(data,2)
```

----------------------------------------------------
FUNCTION ------------ PLEASE HELP

Trying to get this function to take my data file, seperate by status, reverse order of dates, loop through dates backwards, get active cases value, and create a new row in my dataframe.

Getting an error on line 308 for passing the date.

```{r}

```


```{r}
p <- ggplot(totals[totals$Country != "Global",], aes(reorder(Status, Total), Total, fill = Country))+
  geom_bar(stat="identity")+ coord_flip()+ xlab("Status of Case")+ ylab("Total Cases")+ ggtitle("Number of Cases by Country")
fig <- ggplotly(p)
fig
```

The above graph shows the variance of cases by type, and by country. We can see some pretty interesting things here. For example, Italy has the highest deaths and active cases, yet they are only second in the number of confirmed cases. This could mean two things:
1. Italy may not have been testing enough people, which would bring their death-rate up.
2. Italys' healthcare system is getting overrun, and their deaths are about to skyrocket.

```{r}
a <- ggplot(data_c[data_c$Country == countries,], aes(Date, Value, color = Country))+
  geom_line()+ geom_point(aes(size = Rate_Of_Change),show.legend = FALSE)+
          scale_x_date(labels=date_format("%b/%d"), date_breaks = "1 week", date_minor_breaks = "1 day")+
            xlab("") + ylab("Count")+
                ggtitle("Timeline For Number of CONFIRMED Cases")
figa <- ggplotly(a)
figa
```

```{r}
a <- ggplot(data_c[data_c$Country == countries,], aes(Date, Value, color = Country))+
  geom_line()+ geom_point(aes(size = Rate_Of_Change),show.legend = FALSE)+
          scale_x_date(labels=date_format("%b/%d"), date_breaks = "1 week", date_minor_breaks = "1 day")+
            xlab("") + ylab("# of Cases")+
                ggtitle("Timeline For Number of CONFIRMED Cases")
figa <- ggplotly(a)

b <- ggplot(data[data$Status == "Confirmed" & data$Country == countries,], aes(Date, Rate_Of_Change, color = Country))+
  geom_point(aes(size = Value),show.legend = FALSE)+ scale_x_date(labels=date_format("%b/%d"), date_breaks = "1 week", date_minor_breaks = "1 day")+ geom_line()+
    xlab("") + ylab("Growth Rate")+
        ggtitle("The Rate of Change Per Day for CONFIRMED Cases")
figb <- ggplotly(b)
```

```{r}
figa
figb
```

# Rates of Change
For the above graphs, you can click on the Countries in the legend to select or de-select. Size of the dot correlates the the Total number of cases.

There are two country matchup I want to look at:
* US vs Italy
* US vs China

Theres a couple things to keep in mind when looking at the rate of change.
1. It has more variance in the beginning.
    * As the number of cases get larger, the rate will tend to even out
    
#    **$f(x) = P_0(1+r)^d$**
   
2. The rate is what gets exponentiated for an exponential function, ie.    
    * This means that if our mean rate is higher, our confirmed cases will get exponentially higher with each day. Also, if the rate is negative that means we are starting to lose cases per day.
    
From the graphs we can see that Chinas' number of cases looks like its reached the peak. We might be seeing some negative growth rates in China soon, as its already close with 0.0006 (or 0.06%) change per day. This means that their number of confirmed cases have almost stalled!

Looking at Italy we can see their current growth rate is trending around 0.12, their growth has significantly slowed, most likely due to the shut downs they have been doing. Comparing Italy to the US, we see that the US has over 2x the growth rate of Italy! It seems to be pretty steady around 0.25 (or 25%). This means if there are 1,000 confirmed cases to date then there will be roughly 250 new cases tomorrow. 

```{r}
c <- ggplot(data[data$Country=="Italy",], aes(Rate_Of_Change, color = Status, fill = Status))+ geom_density(alpha=0.4)+ ylab("Density")+
  xlab("Rate of Change")+ 
  ggtitle("Density of the Rate of Change for each Status")
fig <- ggplotly(c)
fig
```

This graph clearly shows that the deceased rate of change has not only a higher average, but most of the rates are higher than the rate of confirmed.

This could be due to the fact that the Confirmed cases will always be lagging behind, or it could be due to the beginning of an overburdened healthcare system. It is also a possiblility that since our total Deaths are still low, we don't have enough data to get a good rate, and it is still in the beginning phase, where the rates tend to be higher.

---------------------------------------------

quickmath
```{r}
total_us_confirmed = max(data_c[data_c$Country== "US", "Value"])

total_us_dead = max(data_d[data_d$Country== "US","Value"])

total_us_dead/total_us_confirmed
total_us_confirmed
total_us_dead

data[data$Country== "US" & data$Status == "Deceased","Value"]
```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

-----------------------------------

```{r}
get_active <- function(data, country){
  #delete extra columns for now, function would be callled before they are added
  data = data[,c("Status","Value","Date","Country")]
  
  
  #get status seperated(need to get efficient ordering)
  data1 = extract_status(data, "Confirmed")
  data2 = extract_status(data, "Recovered")
  data3 = extract_status(data, "Deceased")
  
  #reverse order of dates
  data1 = data1[rev(order(as.Date(data1$Date, format   = "%Y-%m-%d"))),]
  data2 = data2[rev(order(as.Date(data2$Date, format   = "%Y-%m-%d"))),]
  data3 = data3[rev(order(as.Date(data3$Date, format   = "%Y-%m-%d"))),]
  
  #loop through dates
  for (d in data1$Date){
      con = data1[data1$Date == d,"Value"] #confirmed
      rec = data2[data2$Date == d,"Value"] #recovered
      dea = data3[data3$Date == d,"Value"] #deceased
      act = con - rec - dea
      
      print(d)
      #trying to add a new row, date is throwing it off
      data1 = rbind(data1, c("Active", act, as.Date.numeric(d,origin = "1970-01-01"), country))
      
  }
  #reassign, should be deleting a few rows
  data = data1
  return(data)
}
#calling function
active <- get_active(data_us,"US")
```

```{r}
#Convert to pdf

library("rmarkdown")
render("Covid-19_Analysis.Rmd", "pdf_document")
```